Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R

### 2017-04-20 meehan_et_al_inla_nmix_jss/article.tex
Small fix in my name
More fine tuning

%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Timothy D. Meehan\\National Audubon Society \And
        Nicole L. Michel\\National Audubon Society \And
        H\"{a}vard Rue\\King Abdulla University\\of Science and Technology}
\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}

@@

$$N_i \sim Poisson(\lambda) \qquad and \qquad  Y_{i,j} | N_i \sim binomial(N_i, p).$$

$\lambda$ is commonly modeled as a log-linear function of site covariates, as $log(\lambda_i) = \beta_0 + \beta_1 * x_i$. Similarly, $p$ is commonly modeled as $logit(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-survey covariates.

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.  In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim Poisson(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim binomial(N_{i,k}, p_{i,k})$.  Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates.  There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}.  Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.  We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for a nice overview.

@@

\begin{Code}
data4sim <- function(n.sites = 72,   # number study sites
                     n.surveys = 3,  # short term replicates
                     n.years = 9,    # number years
                     lam.b0 = 2.0,   # intercept for log lambda
                     lam.b1 = 2.0,   # slope for log lambda, covariate 1
                     lam.b2 = -3.0,  # slope for log lambda, covariate 2
                     lam.b4 = 1.0,   # slope for log lambda, year
                     p.b0 = 1.0,     # intercept for logit p
                     p.b1 = -2.0,    # slope for logit p, covariate 1
                     p.b3 = 1.0,     # slope for logit p covariate 3
                     disp.size = 3.0 # size of the overdisperison
                     ){
  # setup
  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
  N.tr <- array(dim = c(n.sites, n.years))           # array for true abund
  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # array for ex1 counts
  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # array for ex2 counts
  # abundance covariate values
  x.lam.1 <- array(as.numeric(scale(runif(n=n.sites, -0.5, 0.5), scale=F)),
                   dim=c(n.sites, n.years))
  x.lam.2 <- array(as.numeric(scale(runif(n=n.sites, -0.5, 0.5), scale=F)),
                   dim=c(n.sites, n.years))
  yrs <- 1:n.years; yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
  yr <- array(rep(yrs, each=n.sites), dim=c(n.sites, n.years))
  # fill abundance array
  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
  for(i in 1:n.sites){
    for(k in 1:n.years){
    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
  }}
  # detection covariate values
  x.p.1 <- array(x.lam.1[,1], dim=c(n.sites,n.surveys,n.years))
  x.p.3 <- array(as.numeric(scale(runif(n=n.sites*n.surveys*n.years, -0.5, 0.5),
                                  scale=F)), dim=c(n.sites,n.surveys,n.years))
  # average x.p.3 per site-year
  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
  out1 <- c()
  for(k in 1:n.years){
   chunk1 <- x.p.3.mean[,k]
   chunk2 <- rep(chunk1, n.surveys)
   out1 <- c(out1, chunk2)
  }
  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
  # fill count array with site-yr x.p.3
  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
  for (i in 1:n.sites){
    for (k in 1:n.years){
      for (j in 1:n.surveys){
        y1[i,j,k] <- rbinom(1, size=N.tr[i,k], prob=p.tr1[i,j,k])
  }}}
  # fill count array with site-surv-yr x.p.3
  p.tr2 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3)
  for (i in 1:n.sites){
    for (k in 1:n.years){
      for (j in 1:n.surveys){
        y2[i,j,k] <- rbinom(1, size=N.tr[i,k], prob=p.tr2[i,j,k])
  }}}
  # return data
  return(list(n.sites=n.sites, n.surveys=n.surveys, n.years=n.years,
              x.p.1=x.p.1[,1,1], x.p.3=x.p.3, x.p.3.mean=x.p.3.mean,
              x.p.3.arr=x.p.3.arr, x.lam.1=x.lam.1[,1],
              x.lam.2=x.lam.2[,1], yr=yr[1,], y1=y1, y2=y2,
              lam.tr=lam.tr, N.tr=N.tr))
} #end function

# create reproducible datasets for exercises 1 and 2
set.seed(12345)
sim.data <- data4sim()
\end{Code}

@@

\begin{Code}
jags.model.string <- "
model {
  # priors
  intP ~ dnorm(0, 0.01)       # detection intercept
  bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
  bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
  intLam ~ dnorm(0, 0.01)     # lambda intercept
  bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
  bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
  bYr ~ dnorm(0, 0.01)        # year effect
  overDisEst ~ dunif(0, 5)    # overdispersion size
  # abundance component
  for (k in 1:nYears){
    for (i in 1:nSites){
      N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
      prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
      log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
                          (bYr * yr[k])
  # detection component
      for (j in 1:nSurveys){
        y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
        p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
        lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
      } # close j loop
    } # close i loop
  } # close k loop
} # close model loop
"
}
\end{Code}

@@

\begin{Code}
# parameters to monitor
params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
            "bYr", "overDisEst")
# jags data
jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
             x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
             x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
             nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
             nYears = sim.data$n.years)
# initial values
N.init <- sim.data$y1 # initial count values
N.init[is.na(N.init)] <- 1 # clean up NA's
N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
                     intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
                     bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01),
                     bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
                     overDisEst = runif(1, 0.5, 2.5))
\end{Code}

@@

\begin{Code}
# set run parameters
nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10
# run jags
out.jags <- run.jags(model = jags.model.string, data = jags.data,
                     monitor = params, n.chains = nc, inits = inits,
                     burnin = nb, adapt = na, sample = ni, thin = nt,
                     modules = "glm on", method = "parallel")
summary(out.jags)
\end{Code}

@@

\begin{Code}
# format count data
y.unmk <- sim.data$y1[ , , 1]
for(i in 2:sim.data$n.years){
  y.chunk <- sim.data$y1[ , , i]
  y.unmk <- rbind(y.unmk, y.chunk)
}
# format covariates
x.lam.1.unmk <- rep(sim.data$x.lam.1, sim.data$n.years)
x.lam.2.unmk <- rep(sim.data$x.lam.2, sim.data$n.years)
yr.unmk <- rep(sim.data$yr, each = sim.data$n.sites)
x.p.1.unmk <- rep(sim.data$x.p.1, sim.data$n.years)
x.p.3.unmk <- c(sim.data$x.p.3.mean)
site.covs.unmk <- data.frame(x.lam.1.unmk, x.lam.2.unmk, yr.unmk,
                             x.p.1.unmk, x.p.3.unmk)
# make unmarked pcount frame
unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
\end{Code}

@@

\begin{Code}
# run unmk model
out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
                   ~ x.lam.1.unmk + x.lam.2.unmk + yr.unmk,
                    data = unmk.data, mixture = "NB")
summary(out.unmk)
\end{Code}

@@

\begin{Code}
# format count data
y.inla <- sim.data$y1[ , , 1]
for(i in 2:sim.data$n.years){
  y.chunk <- sim.data$y1[ , , i]
  y.inla <- rbind(y.inla, y.chunk)
}
# format covariates
x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
x.p.3.inla <- c(sim.data$x.p.3.mean)
# make inla.mdata object
counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, x.lam.2.inla, yr.inla)
\end{Code}

@@

\begin{Code}
# run inla model
out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
         data = list(counts.and.count.covs = counts.and.count.covs,
                     x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
         family = "nmixnb",
         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
                              prec.intercept = 0.01),
         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
                                          theta2 = list(param = c(0, 0.01)),
                                          theta3 = list(param = c(0, 0.01)),
                                          theta4 = list(param = c(0, 0.01)),
                                          theta5 = list(prior = "flat",
                                                        param = numeric()))))
summary(out.inla)
\end{Code}

@@

\begin{Code}
# get real mallard data from unmarked package
data(mallard)
# format unmarked data
mallard.y.unmk <- mallard.y
mallard.site.unmk <- mallard.site
mallard.obs.unmk <- mallard.obs
mallardUMF <- unmarkedFramePCount(mallard.y.unmk, siteCovs = mallard.site.unmk,
                                  obsCovs = mallard.obs.unmk)
# run unmarked model
mallard.out.unmk <- pcount(~ ivel+ date + I(date^2) ~ length + elev + forest,
                           mixture = "NB", mallardUMF, K=30)
summary(mallard.out.unmk)
\end{Code}

@@

\begin{Code}
# format inla data
mallard.y.inla <- mallard.y
mallard.length.inla <- mallard.site[,2]
mallard.elev.inla <- mallard.site[,1]
mallard.forest.inla <- mallard.site[,3]
mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T) # average per site
mallard.ivel.inla[is.na(mallard.ivel.inla)] <- mean(mallard.ivel.inla, na.rm=T)
mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T) # average per site
mallard.date2.inla <- mallard.date.inla^2
# make inla.mdata object
counts.and.count.covs <- inla.mdata(mallard.y.inla, 1, mallard.length.inla,
                                    mallard.elev.inla, mallard.forest.inla)
# run inla model
mallard.out.inla <- inla(counts.and.count.covs ~ 1 + mallard.ivel.inla +
                           mallard.date.inla + mallard.date2.inla,
         data = list(counts.and.count.covs = counts.and.count.covs,
                     mallard.ivel.inla = mallard.ivel.inla,
                     mallard.date.inla = mallard.date.inla,
                     mallard.date2.inla = mallard.date2.inla),
         family = "nmixnb",
         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
                              prec.intercept = 0.01),
         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
                                          theta2 = list(param = c(0, 0.01)),
                                          theta3 = list(param = c(0, 0.01)),
                                          theta4 = list(param = c(0, 0.01)),
                                          theta5 = list(prior = "flat",
                                                        param = numeric()))))
summary(mallard.out.inla)
\end{Code}

@@

%% appendix
\appendix
\section{Recursive computations of the "nmix" likelihood}
The likelihood for the simplest case is

\begin{displaymath}
    \text{Probability}(y) = \sum_{n = y}^{\infty}
    \text{Poisson}(n ; \lambda) \;\times\; \text{binomial}(y;  n, p)
\end{displaymath}

where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{binomial}(y; n, p)$ is the density for the binomial distribution
with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$. Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we
can easily evaluate it using a recursive algorithm that is both faster and more numerical stable. The same idea is also applicable to the negative binomial case, and the case where we have replicated observations of the same $n$. We leave it to the reader to derive these straight forward extentions.

The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,

\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}

and

\begin{displaymath}
    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}

and then also for the Poisson-binomial product

\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}

If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as

\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}

The log-liklihood can then evaluated using the following simple \proglang{R}-code

\begin{verbatim}
fac = 1; ff = lambda * (1-p)
for (i in (n.max - y):1) fac = 1 + fac * ff / i
log.L = dpois(y, lambda, log=TRUE) +
        dbinom(y, y, p, log=TRUE) + log(fac)
\end{verbatim}

>>>>

%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Timothy D. Meehan\\National\\Audubon Society \And
        Nicole L. Michel\\National\\Audubon Society \And
        H\"{a}vard Rue\\King Abdulla University\\of Science and Technology}
\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}

@@

$$N_i \sim \text{Poisson}(\lambda) \qquad \text{and} \qquad  Y_{i,j} | N_i \sim \text{binomial}(N_i, p).$$

$\lambda$ is commonly modeled as a log-linear function of site covariates, as $\text{log}(\lambda_i) = \beta_0 + \beta_1 * x_i$. Similarly, $p$ is commonly modeled as $\text{logit}(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-survey covariates.

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.  In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{binomial}(N_{i,k}, p_{i,k})$.  Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates.  There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}.  Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.  We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for a nice overview.

@@

\begin{Code}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # short term replicates
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.b4 = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{Code}

@@

\begin{Code}
R> jags.model.string <- "
+    model {
+    # priors
+    intP ~ dnorm(0, 0.01)       # detection intercept
+    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
+    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
+    intLam ~ dnorm(0, 0.01)     # lambda intercept
+    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
+    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
+    bYr ~ dnorm(0, 0.01)        # year effect
+    overDisEst ~ dunif(0, 5)    # overdispersion size
+    # abundance component
+    for (k in 1:nYears){
+      for (i in 1:nSites){
+        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
+        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
+        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
+                          (bYr * yr[k])
+    # detection component
+        for (j in 1:nSurveys){
+          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
+          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
+          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
+        } # close j loop
+      } # close i loop
+    } # close k loop
+  } # close model loop
+  "
\end{Code}

@@

\begin{Code}

R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
+              "bYr", "overDisEst") # parameters to monitor
R> jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
+               x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
+               x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
+               nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
+               nYears = sim.data$n.years)
R> N.init <- sim.data$y1 # initial count values
R> N.init[is.na(N.init)] <- 1 # clean up NA's
R> N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
R> inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
+                      intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
+                      bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01),
+                      bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
+                      overDisEst = runif(1, 0.5, 2.5))
\end{Code}

@@

\begin{Code}
R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # set run parameters
R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
+                       monitor = params, n.chains = nc, inits = inits,
+                       burnin = nb, adapt = na, sample = ni, thin = nt,
+                       modules = "glm on", method = "parallel")
\end{Code}

@@

\begin{Code}
R> y.unmk <- sim.data$y1[ , , 1]
R> for(i in 2:sim.data$n.years){
+    y.chunk <- sim.data$y1[ , , i]
+    y.unmk <- rbind(y.unmk, y.chunk)
+  } 
R> x.lam.1.unmk <- rep(sim.data$x.lam.1, sim.data$n.years)
R> x.lam.2.unmk <- rep(sim.data$x.lam.2, sim.data$n.years)
R> yr.unmk <- rep(sim.data$yr, each = sim.data$n.sites)
R> x.p.1.unmk <- rep(sim.data$x.p.1, sim.data$n.years)
R> x.p.3.unmk <- c(sim.data$x.p.3.mean)
R> site.covs.unmk <- data.frame(x.lam.1.unmk, x.lam.2.unmk, yr.unmk,
+                               x.p.1.unmk, x.p.3.unmk)
R> unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
\end{Code}

@@

\begin{Code}
R> out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
+                     ~ x.lam.1.unmk + x.lam.2.unmk + yr.unmk,
+                     data = unmk.data, mixture = "NB")
\end{Code}

@@

\begin{Code}

R> y.inla <- sim.data$y1[ , , 1]
R> for(i in 2:sim.data$n.years){
+    y.chunk <- sim.data$y1[ , , i]
+    y.inla <- rbind(y.inla, y.chunk)
+  }
R> x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
R> x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
R> yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
R> x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
R> x.p.3.inla <- c(sim.data$x.p.3.mean)
R> counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, 
+                                      x.lam.2.inla, yr.inla)
\end{Code}

@@

\begin{Code}
R> out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
+           data = list(counts.and.count.covs = counts.and.count.covs,
+                       x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
+           family = "nmixnb",
+           control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
+                                prec.intercept = 0.01),
+           control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
+                                              theta2 = list(param = c(0, 0.01)),
+                                              theta3 = list(param = c(0, 0.01)),
+                                              theta4 = list(param = c(0, 0.01)),
+                                              theta5 = list(prior = "flat",
+                                                            param = numeric()))))
\end{Code}

@@

\begin{Code}
R> data("mallard") # get mallard data from unmarked package
R> mallard.y.unmk <- mallard.y # format unmarked data
R> mallard.site.unmk <- mallard.site
R> mallard.obs.unmk <- mallard.obs
R> mallardUMF <- unmarkedFramePCount(mallard.y.unmk, 
+                                    siteCovs = mallard.site.unmk,
+                                    obsCovs = mallard.obs.unmk)
R> mallard.out.unmk <- pcount(~ ivel+ date + I(date^2)
+                             ~ length + elev + forest,
+                             mixture = "NB", mallardUMF, K=30) 
\end{Code}

@@

\begin{Code}
R> mallard.y.inla <- mallard.y # format inla data
R> mallard.length.inla <- mallard.site[,2]
R> mallard.elev.inla <- mallard.site[,1]
R> mallard.forest.inla <- mallard.site[,3]
R> mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T) # site average
R> mallard.ivel.inla[is.na(mallard.ivel.inla)] <- 
+                                       mean(mallard.ivel.inla, na.rm = T)
R> mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T) # site average
R> mallard.date2.inla <- mallard.date.inla^2
R> counts.and.count.covs <- inla.mdata(mallard.y.inla, 1, 
+                                      mallard.length.inla,
+                                      mallard.elev.inla, 
+                                      mallard.forest.inla)
R> mallard.out.inla <- inla(counts.and.count.covs ~ 1 + mallard.ivel.inla +
+                           mallard.date.inla + mallard.date2.inla,
+      data = list(counts.and.count.covs = counts.and.count.covs,
+                           mallard.ivel.inla = mallard.ivel.inla,
+                           mallard.date.inla = mallard.date.inla,
+                           mallard.date2.inla = mallard.date2.inla),
+      family = "nmixnb",
+      control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
+                               prec.intercept = 0.01),
+      control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
+                                    theta2 = list(param = c(0, 0.01)),
+                                    theta3 = list(param = c(0, 0.01)),
+                                    theta4 = list(param = c(0, 0.01)),
+                                    theta5 = list(prior = "flat",
+                                                  param = numeric()))))
\end{Code}

@@

%% appendix
\appendix
\section{Recursive computations of the "nmix" likelihood}
The likelihood for the simplest case is
\begin{displaymath}
    \text{Probability}(y) = \sum_{n = y}^{\infty}
    \text{Poisson}(n ; \lambda) \;\times\; \text{binomial}(y;  n, p)
\end{displaymath}
where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{binomial}(y; n, p)$ is the density for the binomial distribution
with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$. Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we
can easily evaluate it using a recursive algorithm that is both faster and more numerical stable. The same idea is also applicable to the negative binomial case, and the case where we have replicated observations of the same $n$. We leave it to the reader to derive these straight forward extensions.

The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then evaluated using the following simple \proglang{R}-code
\begin{Code}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{Code}
Since this evaluation is recursive in decreasing $n$, we have to chose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.
